---
title: "BIMM 143 Project 3"
output:
  html_document:
    df_print: paged
---

# Effects of different cannabinoids on activated microglia transcription of a few key genes 

### BIMM 143 Project 3: Daniel Naranjo Sampson 

## Scientific Question:

What will be the difference in gene expression between CBD and THC interactions with microglia in their different immune activation states? 

To answer this question, I'll be using a dataset I pulled from a whitepaper on the same topic and performing gene ontology (GO) to make network maps of the genes depicting their interactions and fold changes. Afterwards, since I did way too much in following this one tutorial, I'm gonna also include a snippet of code doing pathway enrichment.

Analysis will involve looking at a couple key genes involved in inflammatory processes including nitric oxide synthase and the CB2 receptor and seeing how their expressions are altered across the five total conditions. 

## Background info on the data 
We're looking at Microarray data for Mus musculus BV-2 microglia cells in different conditions. These immune cells are activated with lipopolysaccharide (LPS), which is a pathogen-associated molecular pattern (PAMP) for bacterial infection. In other words, the microglia applied with LPS shoud react as if there were a pathogen nearby. 

This study basically looks at the differential gene expression between these microglial cells with either THC or CBD added in with or without LPS-activation. 

The main mode of effect for either cannabinoid will be the CB2 receptor, whose function is typically to increase microglia proliferation in anti-inflammatory way where they produce and release less inflammatory factors like TNFα and free radicals, downregulate certain viral-activated chemokine receptors, and increase release of beneficial factors like BDNF. I should also mention that microglia actually produce their own endocannabinoids (anandamide and 2-acylglycerol aka 2-AG), so some of this signaling is autocrine, so maybe using cannabinoids like THC and CBD as agonists will have some of the same effect as the endocannabinoids in that respect. 

For my purposes I'll only analyze the effect on a few key genes involved in different microglia functions mostly related to the immune response by comparing their expressions across the various conditions. Since my data is partitioned into LPS-up and down regulated gene lists, I'll chose three from each list plus the CB2 receptor. 

- LPS-upregulated genes: Nos2 (iNOS), Il-6, and Ccl2 (also known as MCP-1)
- LPS-downregulated genes: CD28, Cxcr4, Dgkg, and Cnr2 (CB2 endocannabinoid receptor)

#### Necessary Packages

Importing packages we need for the code we're running. Quite a few of them but the most important ones have to do with our Gene Ontology (GO) and pathway analysis. org.Mm.eg.db, clusterProfiler, DOSE, rWikiPathways, RCy3 all are imports required for that. data.table, ggplot2, ggnewscale are all packages relevant to data selection and visualization, while dplyr is an all-around useful package for working with large datasets. 

- Org.db are gene databases maintained by Bioconductor for model organisms with org.Mm.eg.db being the database for Mus musculus. 
- clusterProfiler contains most of the gene ontology functions we use here
- DOSE is a parent package to clusterProfiler, since the latter draws a ton of functionality from DOSE. 
- rWikiPathways is a package for pathway enrichment, and towards the end I go through its functionality as well. 
- data.table is the parent package to data.frame. Honestly I think I had to include it to solve a random error that popped up down the line, because I've never had to import it manually before now.
- dplyr is a super useful function for selecting and manipulating dataframes. 

```{r}
library(data.table)
library(org.Mm.eg.db)
library(clusterProfiler)
library(ggplot2)
library(DOSE)
library(dplyr)
library(rWikiPathways)
```

#### Data Import and formatting 

In these next lines here, we import our data from Excel tables as dataframes. On my computer I set my working directory to Downloads; not sure if this will work for you if you don't have access to the original data. Anyways from there I just format the data: adding in column names and such (which I could've done via the header parameter but when I did so, the format of them was simply a bit of a mess). 

```{r}
# Upregulated genes
upreg_genes <- read.csv(file = "FORMATTED_upregulated_genes - Copy - Copy.csv", header=FALSE,stringsAsFactors = FALSE)
colnames(upreg_genes) <- c('geneName','geneDescription','RefSeqID','foldLPS','foldLPS_CBD','foldLPS_THC','foldCBD','foldTHC','log2LPS','log2LPS_CBD','log2LPS_THC','log2CBD','log2THC','CBD_effect2_LPS','THC_effect2_LPS')

# Downregulated genes
downreg_genes <- read.csv(file = "FORMATTED_downreg_genes.csv", header = FALSE, stringsAsFactors = FALSE)
colnames(downreg_genes) <- c('geneName', 'geneDescription', 'RefSeqID','foldLPS','foldLPS_CBD','foldLPS_THC','foldCBD','foldTHC','log2LPS','log2LPS_CBD','log2LPS_THC','log2CBD','log2THC','CBD_effect2_LPS','THC_effect2_LPS')

# Chopping some rows off of downreg_genes because the two empty rows I left in the Excel sheet were annoying me.
downreg_genes <- downreg_genes[-c(1,2),] 
```

We have to use a regular expression here to chop off the version numbers from the RefSeq gene IDs because later GO functions just won't be able to process those. This gsub() function will change a gene ID from something like "NM_013653.1" to "NM_013653" for the Ccl5 gene. 

```{r}
upreg_genes$RefSeqID <- gsub(pattern = "\\.\\d", replacement = "", upreg_genes$RefSeqID)
downreg_genes$RefSeqID <- gsub(pattern = "\\.\\d", replacement = "", downreg_genes$RefSeqID)
```

Now we get to talk about Gene IDs. I've had to learn way too much about these recently, so you get to join me in my suffering. There's a ton of ways to ID any one gene, between the various naming conventions, languages, inconsistent acronym usage, and of course whether we're talking DNA, mRNA or protein, even the specific country we're in --- it's just too much man. So bioinformaticians have made specific codes for each individual gene, and these follow various formats or 'keyTypes' depending on the database being used.

If you're curious about these keyTypes, use this function below to find out how many there are just for mice.

```{r}
keytypes(org.Mm.eg.db)
```


NCBI and the paper I pull the microarray data from use RefSeq formats, and my GO packages require EntrezIDs, so I use and specifically the bitr() function within the clusterProfiler package to convert my gene lists. 

So let's do that. 

```{r}
up.genes.entrez <- clusterProfiler::bitr(upreg_genes$RefSeqID, fromType = "REFSEQ", toType = "ENTREZID", OrgDb = org.Mm.eg.db, drop = FALSE)
down.genes.entrez <- clusterProfiler::bitr(downreg_genes$RefSeqID,  fromType = "REFSEQ", toType = "ENTREZID", OrgDb = org.Mm.eg.db, drop = FALSE)
```

Now we have our dataframes for up and downregulated genes under the LPS-alone condition, as well as genelists in EntrezID formats to work with. 

#### Gene Ontology 

Now it's time for the gene ontology itself. Here, we're using the clusterProfiler package again, but now it's the groupGO() function. This part also uses dependencies within the DOSE package for the geneList dataset. Also I'm sorting the data within each column before doing anything, although I don't remember 100% why that was necessary. 

You can see the data output from this if you run the head() functions down below. 

```{r}
order(up.genes.entrez$ENTREZID, decreasing = TRUE)
order(down.genes.entrez$ENTREZID, decreasing = TRUE)

upreg_ggo <- clusterProfiler::groupGO(gene = up.genes.entrez$ENTREZID, OrgDb = org.Mm.eg.db)

downreg_ggo <- clusterProfiler::groupGO(gene = down.genes.entrez$ENTREZID, OrgDb = org.Mm.eg.db)

head(upreg_ggo)
head(downreg_ggo)
```

##### Visualization

At this point, we have our GO analysis mostly done, but we need to visualize it to be sure of what we have. So first we're going to make some quick barplots for each dataset. They do not show very much detail but they give us a general idea of the gene categories we're looking at. Also due to the way I did the gene ontologies, ie not separating the gene lists further by upreg/downreg in the various conditions, here we only have our original two categories of LPS-upregulated and LPS-downregulated. 

```{r}
barplot(upreg_ggo, showCategory = 100, title = "LPS-Upregulated Gene Ontology")
barplot(downreg_ggo, showCategory = 100,title = "LPS-Downregulated Gene Ontology")
```

We can also make Gene Concept Networks, which are networks of genes and the various categories they might belong to. Here we can do one for all five experimental groups we have: LPS, LPS+CBD, LPS+THC, CBD, and THC. But first we're gonna convert our EntrezIDs to normal gene SYMBOLs using the setReadable() function within DOSE. Symbols are another keyType, however they're a lot more recognizable to anyone whose studied any genes before because they look like the simple acronyms we see in class and in research papers. Eg. Il-6, Stat1 or NFkB.

```{r}
up_ggo <- setReadable(upreg_ggo, 'org.Mm.eg.db', 'ENTREZID')
str(up_ggo)

dn_ggo <- setReadable(downreg_ggo, 'org.Mm.eg.db', 'ENTREZID')
str(dn_ggo)
```

Now we create the objects representing our gene concept networks. It's a lot of chunky code here and it'll take a lot of lines to write this out, but our main parameters are 1) the list of GO terms and 2) the vector of fold changes for each gene. Everything else is just set dressing for the visuals. Speaking of which, note to self: need to figure out how to add titles to these. 

```{r}
# Upreg genes first 
up_genes_LPS <- cnetplot(up_ggo, foldChange = upreg_genes$foldLPS,
                         colorEdge = TRUE,
                         cex_gene = 1,cex_label_gene = .5,
                         cex_label_category = .01,
                         cex_category = .1) + ggtitle("Upreg genes: LPS only")

up_genes_LPS_CBD <- cnetplot(up_ggo, foldChange = upreg_genes$foldLPS_CBD,
                         colorEdge = TRUE,
                         cex_gene = 1,cex_label_gene = .5,
                         cex_label_category = .01,
                         cex_category = .1) + ggtitle("Upreg genes: LPS + CBD")

up_genes_LPS_THC <- cnetplot(up_ggo, foldChange = upreg_genes$foldLPS_THC,
                         colorEdge = TRUE,
                         cex_gene = 1,cex_label_gene = .5,
                         cex_label_category = .01,
                         cex_category = .1) + ggtitle("Upreg genes: LPS + THC")

up_genes_CBD <- cnetplot(up_ggo, foldChange = upreg_genes$foldCBD,
                         colorEdge = TRUE,
                         cex_gene = 1,cex_label_gene = .5,
                         cex_label_category = .01,
                         cex_category = .1) + ggtitle("Upreg genes: CBD only")

up_genes_THC <- cnetplot(up_ggo, foldChange = upreg_genes$foldTHC,
                         colorEdge = TRUE,
                         cex_gene = 1,cex_label_gene = .5,
                         cex_label_category = .01,
                         cex_category = .1) + ggtitle("Upreg genes: THC only")


# Downreg genes
down_genes_LPS <- cnetplot(dn_ggo, foldChange = downreg_genes$foldLPS,
                           colorEdge = TRUE,
                           cex_gene = 1,cex_label_gene = .5,
                           cex_label_category = .01,
                           cex_category = .1) + ggtitle("Downreg genes: LPS only")

down_genes_LPS_CBD <- cnetplot(dn_ggo, foldChange = downreg_genes$foldLPS_CBD,
                             colorEdge = TRUE,
                             cex_gene = 1,cex_label_gene = .5,
                             cex_label_category = .01,
                             cex_category = .1) + ggtitle("Downreg genes: LPS + CBD")

down_genes_LPS_THC <- cnetplot(dn_ggo, foldChange = downreg_genes$foldLPS_THC,
                             colorEdge = TRUE,
                             cex_gene = 1,cex_label_gene = .5,
                             cex_label_category = .01,
                             cex_category = .1) + ggtitle("Downreg genes: LPS + THC")

down_genes_CBD <- cnetplot(dn_ggo, foldChange = downreg_genes$foldCBD,
                           colorEdge = TRUE,
                           cex_gene = 1,cex_label_gene = .5,
                           cex_label_category = .01,
                           cex_category = .1) + ggtitle("Downreg genes: CBD only")

down_genes_THC <- cnetplot(dn_ggo, foldChange = downreg_genes$foldTHC,
                           colorEdge = TRUE,
                           cex_gene = 1,cex_label_gene = .5,
                           cex_label_category = .01,
                           cex_category = .1) + ggtitle("Downreg genes: THC only")

```

We can look at the fruits of our hard labor by simply calling any one of these new cnetplot() objects we've created. These plots are honestly super cool so I recommend you go through each one individually. 

```{r}
up_genes_LPS
up_genes_LPS_CBD
up_genes_LPS_THC
up_genes_CBD
up_genes_THC

down_genes_LPS
down_genes_LPS_CBD
down_genes_LPS_THC
down_genes_CBD
down_genes_THC
```

### Results and Analysis

So we got all these fine-looking plots right here. Lots of good stuff to be analyzed here; but not by me. Not today. I'll just stick to my original objective and comparing the expression profiles of our seven target genes. The objective now is then to pull those values from our up_ggo and down_ggo results lists. For that, we'll make a some vectors of the genes we're looking for and use a dplyr filter on our original gene dataframes to retrieve their data... 

Or I would if I could on both gene lists. Upreg_genes is coming up with "Error: Can't transform a data frame with duplicate names" when there literally no duplicate rownames anywhere (I checked manually)... so after hours of testing other solutions I have to hardcode this in. This  might take the cake for the stupidest and most inconvenient error I've ever found in my 5 years of programming. 

```{r}
geneName <- c("Nos2", "Il6", "Ccl2")
RefSeqID <- c("NM_010927" ,"NM_031168","NM_011333")
foldLPS <- c(upreg_genes[4,65],upreg_genes[4,6],upreg_genes[4,23])
foldLPS_CBD <- c(upreg_genes[5,65],upreg_genes[5,6],upreg_genes[5,23])
foldLPS_THC <- c(upreg_genes[6,65],upreg_genes[6,6],upreg_genes[6,23])
foldCBD <- c(upreg_genes[7,65],upreg_genes[7,6],upreg_genes[7,23])
foldTHC <- c(upreg_genes[8,65],upreg_genes[8,6],upreg_genes[8,23])

up_target_jeans <- data.frame(geneName,RefSeqID,foldLPS,foldLPS_CBD,foldLPS_THC,foldCBD,foldTHC)

up_target_jeans 

dn_target_jeans <- filter(downreg_genes, geneName %in% c("Cd28","Cxcr4", "Dgkg", "Cnr2")) 
dn_target_jeans <- subset(dn_target_jeans, select=-c(2,9:18))
dn_target_jeans$RefSeqID <- gsub(pattern = "\\.\\d", replacement = "", dn_target_jeans$RefSeqID)

target_genes <- rbind(up_target_jeans, dn_target_jeans)
```

Now that we have a new dataframe with only our genes of interest, we can just do a quick comparison across all the genes listed 
```{r}
target_genes_entrez <- clusterProfiler::bitr(target_genes$RefSeqID, fromType = "REFSEQ", toType = "ENTREZID", OrgDb = org.Mm.eg.db, drop = FALSE)
target_genes <- cbind(target_genes,target_genes_entrez)
order(target_genes$ENTREZID, decreasing = TRUE)

final_ggo <- clusterProfiler::groupGO(gene = target_genes$ENTREZID, OrgDb = org.Mm.eg.db)
final_ggo@readable = TRUE

barplot(final_ggo) + ggtitle("Gene Ontologies of our Final 7 genes")

```

A more indepth analysis can be done by reading into the table values per gene. 
- Nos2 is the inducible nitric oxide synthase and nitric oxide is an anti-inflammatory molecule at low levels. Looking at its fold changes, it's highest in the LPS-only condition and lowest in both CBD and THC conditions.
- Il-6 is a pro-inflammatory interleukin, and it's highest in the LPS-only condition and lowest in both CBD and THC conditions.
- Ccl2 is a tiny molecule involved in recruiting macrophages and microglia. t's highest in the LPS-only condition and also in the LPS+THC condition and lowest in the CBD-only condition.
- Cnr2 is the CB2 receptor and it's highest in the THC and CBD conditions and lowest in the LPS condition. (but negative fold changes across all conditions)
- Dgkg converts diacylglycerol to a small phospholipid that serves as a pro-inflammatory signaling molecule. It's highest in the CBD-only and lowest in the LPS conditon. 
- CD28 is a proinflammatory receptor and it's highest (with a positive foldchange) in the CBD condition and lowest in the LPS condition. 
- Cxcr4 is a pro-inflammatory chemokine that's highest in the CBD condition with a tiny increase of 0.4 and lowest in the LPS condition. 

Overall the different genes have very unique profiles that aren't predictable based just on their pro or anti-inflammatory natures. CBD and THC both seem to have an anti-inflammatory effect on most of them but for the CD28 and Cxcr4 genes, CBD seems to have a pro-inflammatory nature to it, possibly reflecting the eCB's (endocannabinoid system) original function in microglia to recruit other microglia in an anti-inflammatory fashion. One paper I read suggested CBD is a more 'natural' ligand with a binding profile more similar to anadamide and 2-AG, so this would make perfect sense if true. 

### Playing around with Pathway Analysis in rWikiPathways

Technically my Project 3 is done at the visualization and analysis steps, but I'll take this a bit further before I conclude because the tutorial I followed went a few steps further and I already have the code worked out so might as well present it right? 

First we're working with the rWikiPathways package for pathway enrichment. Step 1 is to download the pathway archive for Mus musculus, then collecting vectors of genes and names from the pathway archive. Then we're doing the pathway enrichment on this, but limiting to just the upregulated genes from here on out because this notebook is long enough as it is. 

```{r}
# --- Download Pathway Archive for Mus musculus

wp.mm.gmt <- rWikiPathways::downloadPathwayArchive(organism = "Mus musculus", format = "gmt")
wp2gene <- readPathwayGMT(wp.mm.gmt)

# --- Select vectors of genes and names 
wpid2gene <- wp2gene %>% dplyr::select(wpid,gene)
wpid2name <- wp2gene %>% dplyr::select(wpid,name)

# --- Pathway Enrichment based on Entrez ID 
# ------- Upreg genes

ewp.up.lps <- clusterProfiler::enricher(up.genes.entrez$ENTREZID,
                                        TERM2GENE = wpid2gene, 
                                        TERM2NAME = wpid2name)

ewp.up.lps <- DOSE::setReadable(ewp.up.lps, org.Mm.eg.db, keyType = 'ENTREZID')

head(ewp.up.lps@result)
```

Now we get even more visuals from our pathway enrichment using a simple barplot. We could use ggplot here as well if we wanted by making ewp.up.lps into a dataframe with as.data.frame(). This gives us a bit more detail as well since GO only looks at one category per gene and pathway analysis looks at all pathways a particular gene is involved in. 

```{r}
# ------- Visualize with barplot
barplot(ewp.up.lps, title = "LPS-Upregulated Pathways")
```

After that, we can do whatever we want with this massive, multiple hundred-kilobyte object of tables, geneIDs, p-values and all that. So here what I'm doing is showing how to select the results dataframe, and showing an example of how to find a pathway and a URL with more info about that pathway. We're using TNF signaling for an example, and the pathway URL it'll pull up will be about Spinal Cord Injury in Mus musculus.

```{r}
# Messing with rWikiPathways a bit 
ewp.up.lps@result # Select the results dataframe from ewp.up.lps


# Using TNF signaling as an example
tnf.pathways <- findPathwayNamesByText("TNF")
tnf.pathways <- as.data.frame(mapk.pathways) # Coerce results to dataframe


# Grab our pathway IDs that we collected earlier
holder1 <- grepl(pattern = "Tnf", ewp.up.lps@result$geneID)

tnf.wpids.up.lps <- as.data.frame(ewp.up.lps@result$ID) %>% dplyr::filter(holder1)

# Pull in a vector of URLs about our pathway of interest
url.up.pathways <- getPathwayInfo(tnf.wpids.up.lps)$url 

browseURL(url.up.pathways[1]) # Open a Wikipathways link in browser about this pathway
```

And that's that! Here are some links to the original papers: 
- Data from: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3634783/#!po=1.02041
- Background from: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2828614/ and https://www.frontiersin.org/articles/10.3389/fnmol.2017.00191/full
